---
title: "Modelling Indigenous forests and landscape structure effects on overall zoonotic and vector-borne diseases"
author: "Julia Barreto, Filipa Palmeirin & Paula Prist"
date: "`r Sys.Date()`"
output:
  rmdformats::readthedown:
    highlight: kate
    thumbnails: false
    lightbox: true
    gallery: false
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = F, message = F, warning = F, error= F)
library(here)
library(tidyr)
library(dplyr)
library(skimr)
library(mgcv)
library(bbmle)
library(DHARMa)
library(broom)
library(cowplot)
theme_set(theme_cowplot())
library(ggplot2)
```


# Data Loading and Preprocessing

Load temporal health data and coordinates, merge and clean the datasets

```{r}
incidence_data <- read.csv(here("data", "incidence_data.csv"), header = T)

zoonotic_incidence_data <- incidence_data %>%
  dplyr::select(country, COD, year, IDH, X, Y, zoonotic.incidence, for_PD, 
                for_ED, for_AI, FS_PLAND, tot_IT, NOTackn_IT, acknlgd_IT, 
                FS_noIT, pm25_SUM) %>%
  # Removing NAs from the dataset
  drop_na(zoonotic.incidence)
```

# Data Overview and Distributions

## Summary statistics and structure for dataset
```{r}
(sk <- skim(zoonotic_incidence_data))
```

## Visualize the distribution of the response variables

Log transformed +1 aiming to reach normality
```{r}
hist(log10(zoonotic_incidence_data$zoonotic.incidence+1))
```

# Modelling

## Absence of effect (aka Baseline Model, 'null')
Fit a model of absence of effect with only an intercept and random effects for year and spatial coordinates. This model serves as a baseline to compare against models with predictors. 

```{r}
mzoonotic0 <- gam(log10(zoonotic.incidence+1) ~ 1 + 
                offset(IDH) + s(year, bs = "re") + s(X, Y), 
                data = zoonotic_incidence_data)
# summary(mzoonotic0) ; AIC(mzoonotic0)
```

## Single-Predictor Models

Fit models with single predictors for (1) overall forest cover (FS_PLAND), (2) indigenous territory (tot_IT), (3) forest cover outside indigenous territory (FS_noIT), (4) patch density (PD), (5) edge density (ED).
Here, we first tested overall forest cover, IT and forest cover outside IT with both linear and polynomial (quadratic) terms to determine which has the best fit.

```{r}

# Polynomial model (second degree) for FS_PLAND
mzoonotic_FS2 <- gam(log10(zoonotic.incidence+1) ~ poly(scale(FS_PLAND), 2) + offset(IDH) +  s(year, bs = "re") + s(X, Y),  
                data = zoonotic_incidence_data)
summary(mzoonotic_FS2); AIC(mzoonotic_FS2) 

# Polynomial model (second degree) for tot_IT
mzoonotic_IT2 <- gam(log10(zoonotic.incidence+1) ~ poly(scale(tot_IT), 2) + offset(IDH) +  s(year, bs = "re") + s(X, Y),  
                data = zoonotic_incidence_data)
summary(mzoonotic_IT2); AIC(mzoonotic_IT2)  # Display summary and AIC for the polynomial model

# Model for forest edge density (for_ED)
mzoonotic_ED <- gam(log10(zoonotic.incidence+1) ~ scale(for_ED) +offset(IDH) + s(year, bs = "re") + s(X, Y), 
                data = zoonotic_incidence_data)
# summary(mzoonotic_ED); AIC(mzoonotic_ED)  # Display summary and AIC for for_ED model

# Model for patch density (for_PD)
mzoonotic_PD <- gam(log10(zoonotic.incidence+1) ~ scale(for_PD) + offset(IDH) + s(year, bs = "re") + s(X, Y),  
                data = zoonotic_incidence_data)
# summary(mzoonotic_PD); AIC(mzoonotic_PD)  # Display summary and AIC for for_PD model
```

## Double additive Models I: indigenous territories and forest cover outside IT
Build additive models combining Indigenous Territories (tot_IT) and forest cover outside Indigenous Territories (FS_noIT)
```{r}
# Additive model with polynomial terms for tot_IT and FS_noIT
mzoonotic_ITFSno <- gam(log10(zoonotic.incidence+1) ~ poly(scale(tot_IT), 2) + poly(scale(FS_noIT), 2) + offset(IDH) + s(year, bs = "re") + s(X, Y),  
                data = zoonotic_incidence_data)
# summary(mzoonotic_ITFSno); AIC(mzoonotic_ITFSno)  # Display summary and AIC for the model
```

### Double additive models II: overall forest cover and fragmentation
Build models combining overall forest cover (FS_PLAND) and fragmentation (for_ED and for_PD)
```{r}
# Additive model combining FS_PLAND and for_ED
mzoonotic_FSED <- gam(log10(zoonotic.incidence+1) ~  poly(scale(FS_PLAND), 2) + scale(for_ED) + offset(IDH) + s(year, bs = "re") + s(X, Y),  
                data = zoonotic_incidence_data)
# summary(mzoonotic_FSED); AIC(mzoonotic_FSED)  # Display summary and AIC for the model

# Additive model combining FS_PLAND and for_PD
mzoonotic_FSPD <- gam(log10(zoonotic.incidence+1) ~ poly(scale(FS_PLAND), 2) + scale(for_PD) + offset(IDH) + s(year, bs = "re") + s(X, Y),  
                data = zoonotic_incidence_data)
# summary(mzoonotic_FSPD); AIC(mzoonotic_FSPD)  # Display summary and AIC for the model
```

### Double additive models III: indigenous territories and fragmentation
```{r}
mzoonotic_TIED <- gam(log10(zoonotic.incidence+1) ~ poly(scale(tot_IT),2) + scale(for_ED) + offset(IDH) + s(year, bs = "re") + s(X, Y),
                data = zoonotic_incidence_data)
# summary(mzoonotic_TIED); AIC(mzoonotic_TIED)

mzoonotic_TIPD <- gam(log10(zoonotic.incidence+1) ~ poly(scale(tot_IT),2) + scale(for_PD) + offset(IDH) + s(year, bs = "re") + s(X, Y),
                data = zoonotic_incidence_data)
# summary(mzoonotic_TIPD); AIC(mzoonotic_TIPD)
```

### Tripple additives: indigenous territories + forest cover outside IT + fragmentation
```{r}
mzoonotic_ITFSED <- gam(log10(zoonotic.incidence+1) ~ poly(scale(tot_IT),2) +  poly(scale(FS_noIT),2) + scale(for_ED) + offset(IDH) + s(year, bs = "re") + s(X, Y), 
                data = zoonotic_incidence_data)
# summary(mzoonotic_ITFSED); AIC(mzoonotic_ITFSED)

mzoonotic_ITFSPD <- gam(log10(zoonotic.incidence+1) ~ poly(scale(tot_IT),2) + poly(scale(FS_noIT),2) + scale(for_PD) + offset(IDH) + s(year, bs = "re") + s(X, Y), 
                data = zoonotic_incidence_data)
# summary(mzoonotic_ITFSPD); AIC(mzoonotic_ITFSPD) 
```

## Interactive Models I: indigenous territories * forest cover outside IT
```{r}
# interativos IT * forest (FS_noIT)
mzoonotic_ITFSno1i <- gam(log10(zoonotic.incidence+1) ~ poly(scale(tot_IT),2) * scale(FS_noIT) + offset(IDH) + s(year, bs = "re") + s(X, Y),  
                data = zoonotic_incidence_data)
# summary(mzoonotic_ITFSno1i); AIC(mzoonotic_ITFSno1i) 

mzoonotic_IT1Fsnoi <- gam(log10(zoonotic.incidence+1) ~ scale(tot_IT) * poly(scale(FS_noIT),2) + offset(IDH) + s(year, bs = "re") + s(X, Y),  
                data = zoonotic_incidence_data)
# summary(mzoonotic_IT1Fsnoi); AIC(mzoonotic_IT1Fsnoi) 
```

### Interactive Models II: overall forest cover * fragmentation
```{r}
## interactive landscape structure
mzoonotic_FSEDi <- gam(log10(zoonotic.incidence+1) ~ poly(scale(FS_PLAND),2) * scale(for_ED) + offset(IDH) + s(year, bs = "re") + s(X, Y),  
                data = zoonotic_incidence_data)
# summary(mzoonotic_FSEDi); AIC(mzoonotic_FSEDi) 

mzoonotic_FSPDi <- gam(log10(zoonotic.incidence+1) ~ poly(scale(FS_PLAND),2) * scale(for_PD) + offset(IDH) + s(year, bs = "re") + s(X, Y),  
                data = zoonotic_incidence_data)
# summary(mzoonotic_FSPDi); AIC(mzoonotic_FSPDi) 
```

## Interactive Models III: indigenous territories * fragmentation
```{r}
mzoonotic_TIEDi <- gam(log10(zoonotic.incidence+1) ~ poly(scale(tot_IT),2) * scale(for_ED) + offset(IDH) + s(year, bs = "re") + s(X, Y),  
                data = zoonotic_incidence_data)
# summary(mzoonotic_TIEDi); AIC(mzoonotic_TIEDi) 

mzoonotic_TIPDi <- gam(log10(zoonotic.incidence+1) ~  poly(scale(tot_IT),2) * scale(for_PD) + offset(IDH) + s(year, bs = "re") + s(X, Y),  
                data = zoonotic_incidence_data)
# summary(mzoonotic_TIPDi); AIC(mzoonotic_TIPDi) 
```

### Additive and interative model but keeping interaction with only two varibles: 

#### indigenous territories + ( forest cover outside IT * fragmentation )
```{r}
mzoonotic_ITiFSED <- gam(log10(zoonotic.incidence+1) ~ poly(scale(tot_IT),2) + poly(scale(FS_noIT),2) * scale(for_ED) +
                      offset(IDH) + s(year, bs = "re") + s(X, Y), 
                data = zoonotic_incidence_data)
# summary(mzoonotic_ITiFSED); AIC(mzoonotic_ITiFSED)

mzoonotic_ITiFSPD <- gam(log10(zoonotic.incidence+1) ~ poly(scale(tot_IT),2) + poly(scale(FS_noIT),2) * scale(for_PD) +
                      offset(IDH) + s(year, bs = "re") + s(X, Y),  
                data = zoonotic_incidence_data)
# summary(mzoonotic_ITiFSPD); AIC(mzoonotic_ITiFSPD)
```

#### forest cover outside IT + ( indigenous territories * fragmentation )
```{r}
mzoonotic_FSiTIED <- gam(log10(zoonotic.incidence+1) ~  poly(scale(FS_noIT),2) + poly(scale(tot_IT),2) * scale(for_ED) + offset(IDH) + s(year, bs = "re") + s(X, Y), 
                data = zoonotic_incidence_data)
# summary(mzoonotic_FSiTIED); AIC(mzoonotic_FSiTIED)

mzoonotic_FSiTIPD <- gam(log10(zoonotic.incidence+1) ~ poly(scale(FS_noIT),2) + poly(scale(tot_IT),2) * scale(for_PD) + offset(IDH) + s(year, bs = "re") + s(X, Y), 
                data = zoonotic_incidence_data)
# summary(mzoonotic_FSiTIPD); AIC(mzoonotic_FSiTIPD)
```

#### fragmentation + ( indigenous territories * forest cover outside IT )
```{r}
mzoonotic_EDiFSTI1 <- gam(log10(zoonotic.incidence+1) ~ poly(scale(FS_noIT),2) * scale(tot_IT) + scale(for_ED) + offset(IDH) + s(year, bs = "re") + s(X, Y), 
                data = zoonotic_incidence_data)
 #summary(mzoonotic_EDiFSTI1); AIC(mzoonotic_EDiFSTI1)

mzoonotic_EDiFS1TI <- gam(log10(zoonotic.incidence+1) ~ scale(FS_noIT) * poly(scale(tot_IT),2) + scale(for_ED) + offset(IDH) + s(year, bs = "re") + s(X, Y),
                data = zoonotic_incidence_data)
 #summary(mzoonotic_EDiFS1TI); AIC(mzoonotic_EDiFS1TI)

mzoonotic_PDiFSTI1 <- gam(log10(zoonotic.incidence+1) ~ poly(scale(FS_noIT),2) * scale(tot_IT) + scale(for_PD) + offset(IDH) + s(year, bs = "re") + s(X, Y),  
                data = zoonotic_incidence_data)
# summary(mzoonotic_PDiFSTI1); AIC(mzoonotic_PDiFSTI1)

mzoonotic_PDiFS1TI <- gam(log10(zoonotic.incidence+1) ~ scale(FS_noIT) * poly(scale(tot_IT),2) + scale(for_PD) + offset(IDH) + s(year, bs = "re") + s(X, Y),
                data = zoonotic_incidence_data)
# summary(mzoonotic_PDiFS1TI); AIC(mzoonotic_PDiFS1TI)
```

# Model Selection Using AIC Comparison
Perform model selection by comparing all models based on AIC using AICtab function
```{r}
(mod.sel <- AICtab(
              # Absence of effect (aka Baseline Model, 'null')
              mzoonotic0,
              # Single-Predictor Models
              mzoonotic_FS2,
              mzoonotic_IT2,
              mzoonotic_ED,
              mzoonotic_PD,
              # Double additive Models I: indigenous territories and forest cover outside IT
              mzoonotic_ITFSno,
              # Double additive models II: overall forest cover and fragmentation
              mzoonotic_FSED,
              mzoonotic_FSPD,
              # Double additive models III: indigenous territories and fragmentation
              mzoonotic_TIED,
              mzoonotic_TIPD,
              # Tripple additives: indigenous territories + forest cover outside IT + fragmentation
              mzoonotic_ITFSED,
              mzoonotic_ITFSPD,
              ## Interactive Models I: indigenous territories * forest cover outside IT
              mzoonotic_ITFSno1i,
              mzoonotic_IT1Fsnoi,
              # Interactive Models II: overall forest cover * fragmentation
              mzoonotic_FSEDi,
              mzoonotic_FSPDi,
              # Interactive Models III: indigenous territories * fragmentation
              mzoonotic_TIEDi,
              mzoonotic_TIPDi,
              # Additive and interative model but keeping interaction with only two varibles: 
              # indigenous territories + ( forest cover outside IT * fragmentation )
              mzoonotic_ITiFSED,
              mzoonotic_ITiFSPD,
              # forest cover outside IT + ( indigenous territories * fragmentation )
              mzoonotic_FSiTIED,
              mzoonotic_FSiTIPD,
              # fragmentation + ( indigenous territories * forest cover outside IT )
              mzoonotic_EDiFSTI1,
              mzoonotic_EDiFS1TI,
              mzoonotic_PDiFSTI1,
              mzoonotic_PDiFS1TI,
       base = TRUE, weights = TRUE))  # Include null model and calculate weights
```
# Top selected model output

Summarize the selected model to examine the fitted results
```{r}
summary(mzoonotic_ITiFSED)
```

Tidy up the model summary to get a clean output of the parametric terms
```{r}
tidy(mzoonotic_ITiFSED, parametric = TRUE)  #
```

# Model Diagnostics

```{r}
resid <- simulateResiduals(mzoonotic_ITiFSED)
plot(resid)
```

```{r}
testResiduals(resid, plot = TRUE)
```

Look at selected predictors:
```{r}
plotResiduals(resid, zoonotic_incidence_data$FS_noIT); plotResiduals(resid, zoonotic_incidence_data$tot_IT); plotResiduals(resid, zoonotic_incidence_data$for_ED)
```

# Prediction plot

We originally use orthogonal polynomials (i.e. 'poly(predictor_variable, 2)') for model selection due to their statistical advantages, which often leads to more reliable statistical inference. So model selection was based on orthogonal polynomials (`poly(x, 2)`) for robustness, but for practical visualization purpose, we use regular quadratic terms (`I(x^2)`) for clarity. The overall predicted trends (i.e., the shapes of the response curves) produced by a model using orthogonal polynomials (`poly(x, 2)`) and regular quadratic terms (`I(x^2)`) are generally comparable. Both models capture the quadratic relationships in the data.

```{r}
mzoonotic_ITiFSED_quad <- gam(log10(zoonotic.incidence + 1) ~ 
                                scale(tot_IT) + I(scale(tot_IT)^2) +
                            scale(FS_noIT) * scale(for_ED) + I(scale(FS_noIT)^2) * scale(for_ED) + 
                            offset(IDH) + s(year, bs = "re") + s(X, Y),
                          data = zoonotic_incidence_data)

summary(mzoonotic_ITiFSED_quad)
```

```{r, fig.width= 8, fig.height= 4}
new_data <- expand.grid(
  tot_IT = c(15, 45, 75),  # Three levels of Forest Cover Outside IT
  for_ED = c(10, 25, 50),  # Three fixed levels of Edge Density
  FS_noIT = seq(min(zoonotic_incidence_data$FS_noIT, na.rm = TRUE), max(zoonotic_incidence_data$FS_noIT, na.rm = TRUE), length.out = 100),
  IDH = median(zoonotic_incidence_data$IDH, na.rm = TRUE), 
  year = median(zoonotic_incidence_data$year, na.rm = TRUE),
  X = median(zoonotic_incidence_data$X, na.rm = TRUE),
  Y = median(zoonotic_incidence_data$Y, na.rm = TRUE)
)

# Adjust the filtering based on the specific rules for tot_IT and FS_noIT
new_data <- new_data %>% 
  filter((tot_IT == 15 & FS_noIT <= 85) |
         (tot_IT == 45 & FS_noIT <= 55) |
         (tot_IT == 75 & FS_noIT <= 25))

predictions <- predict(mzoonotic_ITiFSED_quad, newdata = new_data, type = "response", se.fit = TRUE)

# back-transforming predictions
new_data$predicted <- 10^predictions$fit - 1
new_data$conf.low <- 10^(predictions$fit - 1.96 * predictions$se.fit) - 1  # Limite inferior IC 95%
new_data$conf.high <- 10^(predictions$fit + 1.96 * predictions$se.fit) - 1  # Limite superior IC 95%

(zoonotic <- ggplot(new_data, aes(x = FS_noIT, y = predicted, 
                                  linetype = as.factor(for_ED), color = as.factor(for_ED))) +
  geom_line(linewidth = 1) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high, fill = as.factor(for_ED)), 
              linetype= 0, alpha = 0.2) +
  labs(
    x = "Forest cover outside IT (%)",
    y = "Zoonotic/Vector-borne\n diseases' incidence",
    color = "Edge density",
    linetype = "Edge density",
    fill = "Edge density") +
  theme(legend.position = "top",
        text = element_text(size = 16),
        axis.title.y = element_text(size = 16, lineheight = 1.1),
        axis.title.x = element_text(size = 16)) +
  scale_color_manual(values = c("#949E4F", "#D4B35A", "#A9822B")) + 
  scale_fill_manual(values = c("#949E4F", "#D4B35A", "#A9822B")) +
    scale_linetype_manual(values = c("solid", "longdash", "dotted")) +
  facet_wrap(~ tot_IT, labeller = as_labeller(c("15" = "Low IT (15%)", 
                                                "45" = "Median IT (45%)", 
                                                "75" = "High IT (75%)"))))

ggsave(here( "outputs", "figures", "zoonotic_topmodel.png"), width= 9, height= 4)
```
