---
title: "Modelling Indigenous forests and landscape structure effects on Malaria"
author: "Julia Barreto, Filipa Palmeirin & Paula Prist"
date: "`r Sys.Date()`"
output:
  rmdformats::readthedown:
    highlight: kate
    thumbnails: false
    lightbox: true
    gallery: false
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = F, message = F, warning = F, error= F)
library(here)
library(tidyr)
library(dplyr)
library(skimr)
library(mgcv)
library(bbmle)
library(DHARMa)
library(broom)
library(cowplot)
theme_set(theme_cowplot())
library(ggplot2)
```



# Data Loading and Preprocessing

Load temporal health data and coordinates, merge and clean the datasets

```{r}
incidence_data <- read.csv(here("data", "incidence_data.csv"), header = T)

malaria_incidence_data <- incidence_data %>%
  dplyr::select(country, COD, year, IDH, X, Y, malaria.incidence, for_PD, 
                for_ED, for_AI, FS_PLAND, tot_IT, NOTackn_IT, acknlgd_IT, 
                FS_noIT, pm25_SUM) %>%
  # Removing NAs from the dataset
  drop_na(malaria.incidence)
```

# Data Overview and Distributions

## Summary statistics and structure for dataset
```{r}
(sk <- skim(malaria_incidence_data))
```

## Visualize the distribution of the response variables

### Log-transformation of malaria diseases incidence
Log trabnsformed +1 to reach normality
```{r}
hist(log10(malaria_incidence_data$malaria.incidence + 1))
```

# Modelling

We build GAM spatial models contain random effects for year and geographical coordinates. For malaria diseases incidence we use gamma family of distribution and log link function.

## Absence of effect (aka Baseline Model, 'null')
Fit a model of absence of effect with only an intercept and random effects for year and spatial coordinates. This model serves as a baseline to compare against models with predictors. 

```{r}
mmalaria0 <- gam(log10(malaria.incidence+1) ~ 1 + offset(IDH) + s(year, bs = "re") + s(X, Y), 
                data = malaria_incidence_data)
# summary(mmalaria0) ; AIC(mmalaria0)
```

## Single-Predictor Models

Fit models with single predictors for (1) overall forest cover (FS_PLAND), (2) indigenous territory (tot_IT), (3) forest cover outside indigenous territory (FS_noIT), (4) patch density (PD), (5) edge density (ED).
Here, we first tested overall forest cover, IT and forest cover outside IT with both linear and polynomial (quadratic) terms to determine which has the best fit.

```{r}
# Model for forest cover within the entire COD/municipality (FS_PLAND)
# Linear model for FS_PLAND  
mmalaria_FS <- gam(log10(malaria.incidence+1) ~ scale(FS_PLAND) + offset(IDH) + s(year, bs = "re") + s(X, Y), 
                data = malaria_incidence_data)
summary(mmalaria_FS); AIC(mmalaria_FS) # Display summary and AIC for the model

# Polynomial model (second degree) for tot_IT
mmalaria_IT2 <- gam(log10(malaria.incidence+1) ~ poly(scale(tot_IT), 2) + offset(IDH) + s(year, bs = "re") + s(X, Y),  
                data = malaria_incidence_data)
summary(mmalaria_IT2); AIC(mmalaria_IT2)  # Display summary and AIC for the polynomial model

# Model for forest edge density (for_ED)
mmalaria_ED <- gam(log10(malaria.incidence+1) ~ scale(for_ED) +offset(IDH) + s(year, bs = "re") + s(X, Y),  
                data = malaria_incidence_data)
# summary(mmalaria_ED); AIC(mmalaria_ED)  # Display summary and AIC for for_ED model

# Model for patch density (for_PD)
mmalaria_PD <- gam(log10(malaria.incidence+1) ~ scale(for_PD) + offset(IDH) + s(year, bs = "re") + s(X, Y),  
                data = malaria_incidence_data)
# summary(mmalaria_PD); AIC(mmalaria_PD)  # Display summary and AIC for for_PD model
```

## Double additive Models I: indigenous territories and forest cover outside IT
Build additive models combining Indigenous Territories (tot_IT) and forest cover outside Indigenous Territories (FS_noIT)
```{r}
# Additive model with polynomial terms for tot_IT and FS_noIT
mmalaria_ITFSno <- gam(log10(malaria.incidence+1) ~ poly(scale(tot_IT), 2) + poly(scale(FS_noIT), 2) + offset(IDH) + s(year, bs = "re") + s(X, Y),  
                data = malaria_incidence_data)
# summary(mmalaria_ITFSno); AIC(mmalaria_ITFSno)  # Display summary and AIC for the model
```

### Double additive models II: overall forest cover and fragmentation
Build models combining overall forest cover (FS_PLAND) and fragmentation (for_ED and for_PD)
```{r}
# Additive model combining FS_PLAND and for_ED
mmalaria_FSED <- gam(log10(malaria.incidence+1) ~  scale(FS_PLAND)  + scale(for_ED) + offset(IDH) + s(year, bs = "re") + s(X, Y),  
                data = malaria_incidence_data)
# summary(mmalaria_FSED); AIC(mmalaria_FSED)  # Display summary and AIC for the model

# Additive model combining FS_PLAND and for_PD
mmalaria_FSPD <- gam(log10(malaria.incidence+1) ~ scale(FS_PLAND)  + scale(for_PD) + offset(IDH) + s(year, bs = "re") + s(X, Y),
                data = malaria_incidence_data)
# summary(mmalaria_FSPD); AIC(mmalaria_FSPD)  # Display summary and AIC for the model
```

### Double additive models III: indigenous territories and fragmentation
```{r}
mmalaria_TIED <- gam(log10(malaria.incidence+1) ~ poly(scale(tot_IT),2) + scale(for_ED) + offset(IDH) + s(year, bs = "re") + s(X, Y),  
                data = malaria_incidence_data)
# summary(mmalaria_TIED); AIC(mmalaria_TIED)

mmalaria_TIPD <- gam(log10(malaria.incidence+1) ~ poly(scale(tot_IT),2) + scale(for_PD) + offset(IDH) + s(year, bs = "re") + s(X, Y), 
                data = malaria_incidence_data)
# summary(mmalaria_TIPD); AIC(mmalaria_TIPD)
```

### Tripple additives: indigenous territories + forest cover outside IT + fragmentation
```{r}
mmalaria_ITFSED <- gam(log10(malaria.incidence+1) ~ poly(scale(tot_IT),2) +  poly(scale(FS_noIT),2) + scale(for_ED) + offset(IDH) +s(year, bs = "re") + s(X, Y),  
                data = malaria_incidence_data)
# summary(mmalaria_ITFSED); AIC(mmalaria_ITFSED)

mmalaria_ITFSPD <- gam(log10(malaria.incidence+1) ~ poly(scale(tot_IT),2) + poly(scale(FS_noIT),2) + scale(for_PD) + offset(IDH) + s(year, bs = "re") + s(X, Y),  
                data = malaria_incidence_data)
# summary(mmalaria_ITFSPD); AIC(mmalaria_ITFSPD) 
```

## Interactive Models I: indigenous territories * forest cover outside IT
```{r}
# interativos IT * forest (FS_noIT)
mmalaria_ITFSno1i <- gam(log10(malaria.incidence+1) ~ poly(scale(tot_IT),2) * scale(FS_noIT) + offset(IDH) + s(year, bs = "re") + s(X, Y),  
                data = malaria_incidence_data)
# summary(mmalaria_ITFSno1i); AIC(mmalaria_ITFSno1i) 

mmalaria_IT1Fsnoi <- gam(log10(malaria.incidence+1) ~ scale(tot_IT) * poly(scale(FS_noIT),2) + offset(IDH) + s(year, bs = "re") + s(X, Y),  
                data = malaria_incidence_data)
# summary(mmalaria_IT1Fsnoi); AIC(mmalaria_IT1Fsnoi) 
```

### Interactive Models II: overall forest cover * fragmentation
```{r}
## interactive landscape structure
mmalaria_FSEDi <- gam(log10(malaria.incidence+1) ~ scale(FS_PLAND) * scale(for_ED) + offset(IDH) + s(year, bs = "re") + s(X, Y),  
                data = malaria_incidence_data)
# summary(mmalaria_FSEDi); AIC(mmalaria_FSEDi) 

mmalaria_FSPDi <- gam(log10(malaria.incidence+1) ~ scale(FS_PLAND) * scale(for_PD) + offset(IDH) + s(year, bs = "re") + s(X, Y),  
                data = malaria_incidence_data)
# summary(mmalaria_FSPDi); AIC(mmalaria_FSPDi) 
```

## Interactive Models III: indigenous territories * fragmentation
```{r}
mmalaria_TIEDi <- gam(log10(malaria.incidence+1) ~ poly(scale(tot_IT),2) * scale(for_ED) + offset(IDH) + s(year, bs = "re") + s(X, Y),
                data = malaria_incidence_data)
# summary(mmalaria_TIEDi); AIC(mmalaria_TIEDi) 

mmalaria_TIPDi <- gam(log10(malaria.incidence+1) ~  poly(scale(tot_IT),2) * scale(for_PD) + offset(IDH) + s(year, bs = "re") + s(X, Y),
                data = malaria_incidence_data)
# summary(mmalaria_TIPDi); AIC(mmalaria_TIPDi) 
```

### Additive and interative model but keeping interaction with only two varibles: 

#### indigenous territories + ( forest cover outside IT * fragmentation )
```{r}
mmalaria_ITiFSED <- gam(log10(malaria.incidence+1) ~ poly(scale(tot_IT),2) + poly(scale(FS_noIT),2) * scale(for_ED) +
                      offset(IDH) + s(year, bs = "re") + s(X, Y), 
                data = malaria_incidence_data)
# summary(mmalaria_ITiFSED); AIC(mmalaria_ITiFSED)

mmalaria_ITiFSPD <- gam(log10(malaria.incidence+1) ~ poly(scale(tot_IT),2) + poly(scale(FS_noIT),2) * scale(for_PD) +  offset(IDH) + s(year, bs = "re") + s(X, Y), 
                data = malaria_incidence_data)
# summary(mmalaria_ITiFSPD); AIC(mmalaria_ITiFSPD)
```

#### forest cover outside IT + ( indigenous territories * fragmentation )
```{r}
mmalaria_FSiTIED <- gam(log10(malaria.incidence+1) ~  poly(scale(FS_noIT),2) + poly(scale(tot_IT),2) * scale(for_ED) + offset(IDH) + s(year, bs = "re") + s(X, Y), 
                data = malaria_incidence_data) 
# summary(mmalaria_FSiTIED); AIC(mmalaria_FSiTIED)

mmalaria_FSiTIPD <- gam(log10(malaria.incidence+1) ~ poly(scale(FS_noIT),2) + poly(scale(tot_IT),2) * scale(for_PD) + offset(IDH) + s(year, bs = "re") + s(X, Y), 
                data = malaria_incidence_data) 
# summary(mmalaria_FSiTIPD); AIC(mmalaria_FSiTIPD)
```

#### fragmentation + ( indigenous territories * forest cover outside IT )
```{r}
mmalaria_EDiFSTI1 <- gam(log10(malaria.incidence+1) ~ poly(scale(FS_noIT),2) * scale(tot_IT) + scale(for_ED) + offset(IDH) + s(year, bs = "re") + s(X, Y),  
                data = malaria_incidence_data)
 summary(mmalaria_EDiFSTI1); AIC(mmalaria_EDiFSTI1)

mmalaria_EDiFS1TI <- gam(log10(malaria.incidence+1) ~ scale(FS_noIT) * poly(scale(tot_IT),2) + scale(for_ED) + offset(IDH) + s(year, bs = "re") + s(X, Y),  
                data = malaria_incidence_data) 
 #summary(mmalaria_EDiFS1TI); AIC(mmalaria_EDiFS1TI)

mmalaria_PDiFSTI1 <- gam(log10(malaria.incidence+1) ~ poly(scale(FS_noIT),2) * scale(tot_IT) + scale(for_PD) + offset(IDH) + s(year, bs = "re") + s(X, Y),  
                data = malaria_incidence_data)
# summary(mmalaria_PDiFSTI1); AIC(mmalaria_PDiFSTI1)

mmalaria_PDiFS1TI <- gam(log10(malaria.incidence+1) ~ scale(FS_noIT) * poly(scale(tot_IT),2) + scale(for_PD) + offset(IDH) + s(year, bs = "re") + s(X, Y),  
                data = malaria_incidence_data) 
# summary(mmalaria_PDiFS1TI); AIC(mmalaria_PDiFS1TI)
```

# Model Selection Using AIC Comparison
Perform model selection by comparing all models based on AIC using AICtab function
```{r}
(mod.sel <- AICtab(
              # Absence of effect (aka Baseline Model, 'null')
              mmalaria0,
              # Single-Predictor Models
              mmalaria_FS,
              mmalaria_IT2,
              mmalaria_ED,
              mmalaria_PD,
              # Double additive Models I: indigenous territories and forest cover outside IT
              mmalaria_ITFSno,
              # Double additive models II: overall forest cover and fragmentation
              mmalaria_FSED,
              mmalaria_FSPD,
              # Double additive models III: indigenous territories and fragmentation
              mmalaria_TIED,
              mmalaria_TIPD,
              # Tripple additives: indigenous territories + forest cover outside IT + fragmentation
              mmalaria_ITFSED,
              mmalaria_ITFSPD,
              ## Interactive Models I: indigenous territories * forest cover outside IT
              mmalaria_ITFSno1i,
              mmalaria_IT1Fsnoi,
              # Interactive Models II: overall forest cover * fragmentation
              mmalaria_FSEDi,
              mmalaria_FSPDi,
              # Interactive Models III: indigenous territories * fragmentation
              mmalaria_TIEDi,
              mmalaria_TIPDi,
              # Additive and interative model but keeping interaction with only two varibles: 
              # indigenous territories + ( forest cover outside IT * fragmentation )
              mmalaria_ITiFSED,
              mmalaria_ITiFSPD,
              # forest cover outside IT + ( indigenous territories * fragmentation )
              mmalaria_FSiTIED,
              mmalaria_FSiTIPD,
              # fragmentation + ( indigenous territories * forest cover outside IT )
              mmalaria_EDiFSTI1,
              mmalaria_EDiFS1TI,
              mmalaria_PDiFSTI1,
              mmalaria_PDiFS1TI,
       base = TRUE, weights = TRUE))  # Include null model and calculate weights
```
# Top selected model output

Summarize the selected model to examine the fitted results
```{r}
summary(mmalaria_ITiFSPD)
```


Tidy up the model summary to get a clean output of the parametric terms
```{r}
tidy(mmalaria_ITiFSPD, parametric = TRUE)  #
```

# Model Diagnostics

```{r}
resid <- simulateResiduals(mmalaria_ITiFSPD)
plot(resid)
```

```{r}
testResiduals(resid, plot = TRUE)
```

Look at selected predictors:
```{r}
plotResiduals(resid, malaria_incidence_data$FS_noIT); plotResiduals(resid, malaria_incidence_data$tot_IT); plotResiduals(resid, malaria_incidence_data$for_PD)
```

# Prediction plot

We originally use orthogonal polynomials (i.e. 'poly(predictor_variable, 2)') for model selection due to their statistical advantages, which often leads to more reliable statistical inference. So model selection was based on orthogonal polynomials (`poly(x, 2)`) for robustness, but for practical visualization purpose, we use regular quadratic terms (`I(x^2)`) for clarity. The overall predicted trends (i.e., the shapes of the response curves) produced by a model using orthogonal polynomials (`poly(x, 2)`) and regular quadratic terms (`I(x^2)`) are generally comparable. Both models capture the quadratic relationships in the data.

```{r}
mmalaria_ITiFSPD_quad <- gam(log10(malaria.incidence + 1) ~ scale(tot_IT) + I(scale(tot_IT)^2) +
                                     scale(FS_noIT) * scale(for_PD) + I(scale(FS_noIT)^2) * scale(for_PD) + 
                                    offset(IDH) + s(year, bs = "re") + s(X,Y),
                          data = malaria_incidence_data)
```

Plot using new_data:
```{r, fig.width= 8, fig.height= 4}
new_data <- expand.grid(
  tot_IT = c(15, 45, 75),
  for_PD = c(0, 0.8, 1.5),  # Three fixed levels of Edge Density
  FS_noIT = seq(min(malaria_incidence_data$FS_noIT, na.rm = TRUE), max(malaria_incidence_data$FS_noIT, na.rm = TRUE), length.out = 100),
  IDH = median(malaria_incidence_data$IDH, na.rm = TRUE), 
  year = median(malaria_incidence_data$year, na.rm = TRUE),
  X = median(malaria_incidence_data$X, na.rm = TRUE),
  Y = median(malaria_incidence_data$Y, na.rm = TRUE)
)

# Adjust the filtering based on the specific rules for tot_IT and FS_noIT
new_data <- new_data %>% 
  filter((tot_IT == 15 & FS_noIT <= 85) |
         (tot_IT == 45 & FS_noIT <= 55) |
         (tot_IT == 75 & FS_noIT <= 25))

predictions <- predict(mmalaria_ITiFSPD_quad, newdata = new_data, type = "response", se.fit = TRUE)

# back-transforming predictions
new_data$predicted <- 10^predictions$fit - 1
new_data$conf.low <- 10^(predictions$fit - 1.96 * predictions$se.fit) - 1  # Limite inferior IC 95%
new_data$conf.high <- 10^(predictions$fit + 1.96 * predictions$se.fit) - 1  # Limite superior IC 95%

(malaria <- ggplot(new_data, aes(x = FS_noIT, y = predicted, 
                                 linetype = as.factor(for_PD), color = as.factor(for_PD))) +
  geom_line(linewidth = 1) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high, fill = as.factor(for_PD)), 
              linetype = 0, alpha = 0.2) +
  labs(
    x = "Forest cover outside IT (%)",
    y = "Malaria diseases' incidence",
    color = "Patch density",
    linetype = "Patch density",
    fill = "Patch density"
  ) +
  theme(legend.position = "top",
        text = element_text(size = 16),
        axis.title.y = element_text(size = 16, lineheight = 1.1),
        axis.title.x = element_text(size = 16)) +
  scale_color_manual(values = c("#949E4F", "#D4B35A", "#A9822B")) + 
  scale_fill_manual(values = c("#949E4F", "#D4B35A", "#A9822B")) +
    scale_linetype_manual(values = c("solid", "longdash", "dotted")) +
  facet_wrap(~ tot_IT, labeller = as_labeller(c("15" = "Low IT (15%)", 
                                                "45" = "Median IT (45%)", 
                                                "75" = "High IT (75%)"))))

ggsave(here("outputs", "figures", "malaria_topmodel.png"), width= 9, height= 4)
```
