---
title: "Modelling Indigenous forests and landscape structure effects on Chagas incidence"
author: "Julia Barreto, Filipa Palmeirin & Paula Prist"
date: "`r Sys.Date()`"
output:
  rmdformats::readthedown:
    highlight: kate
    thumbnails: false
    lightbox: true
    gallery: false
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = F, message = F, warning = F, error= F)
library(here)
library(tidyr)
library(dplyr)
library(skimr)
library(mgcv)
library(bbmle)
library(DHARMa)
library(broom)
library(cowplot)
theme_set(theme_cowplot())
library(ggplot2)
```


# Data Loading and Preprocessing

Load temporal health data and coordinates, merge and clean the datasets

```{r}
incidence_data <- read.csv(here("data", "incidence_data.csv"), header = T)

chagas.occurrence_data <- incidence_data %>%
  dplyr::select(country, COD, year, IDH, X, Y, chagas.occurrence, for_PD, 
                for_ED, for_AI, FS_PLAND, tot_IT, NOTackn_IT, acknlgd_IT, 
                FS_noIT, pm25_SUM) %>%
  # Removing NAs from the dataset
  drop_na(chagas.occurrence)
```

# Data Overview and Distributions

## Summary statistics and structure for dataset
```{r}
(sk <- skim(chagas.occurrence_data))
```

# Modelling

Generally, we build GAM spatial models contain random effects for year and geographical coordinates. For modelling the presence/absence of chagas.occurrence disease we use the binomial family of distribution and log link function.

## Absence of effect (aka Baseline Model, 'null')
Fit a model of absence of effect with only an intercept and random effects for year and spatial coordinates. This model serves as a baseline to compare against models with predictors. 

```{r}
mchagas.occurrence0 <- gam(chagas.occurrence ~ 1 + offset(IDH) + s(year, bs = "re") + s(X, Y),  
              family= binomial(link = "logit"), data = chagas.occurrence_data) 
# summary(mchagas.occurrence0); AIC(mchagas.occurrence0)
```

## Single-Predictor Models

Fit models with single predictors for (1) overall forest cover (FS_PLAND), (2) indigenous territory (tot_IT), (3) forest cover outside indigenous territory (FS_noIT), (4) patch density (PD), (5) edge density (ED).
Here, we first tested overall forest cover, IT and forest cover outside IT with both linear and polynomial (quadratic) terms to determine which has the best fit.

```{r}
# Model for forest cover within the entire COD/municipality (FS_PLAND)

# Polynomial model (second degree) for FS_PLAND
mchagas.occurrence_FS2 <- gam(chagas.occurrence ~ poly(scale(FS_PLAND), 2) + offset(IDH) + s(year, bs = "re") + s(X, Y), family = binomial(link = "logit"), data = chagas.occurrence_data)
summary(mchagas.occurrence_FS2); AIC(mchagas.occurrence_FS2) 

# Model for Indigenous Territories (tot_IT)
# Linear model for tot_IT   2328.366
mchagas.occurrence_IT <- gam(chagas.occurrence ~ scale(tot_IT) + offset(IDH) + s(year, bs = "re") + s(X,Y), family= binomial(link = "logit"), data = chagas.occurrence_data)
# summary(mchagas.occurrence_IT); AIC(mchagas.occurrence_IT) # Display summary and AIC for the polynomial model

# Polynomial model (second degree) for FS_noIT 
mchagas.occurrence_FSnoIT2 <- gam(chagas.occurrence ~ poly(scale(FS_noIT), 2) + offset(IDH) + s(year, bs = "re") + s(X, Y), family = binomial(link = "logit"), data = chagas.occurrence_data)
summary(mchagas.occurrence_FSnoIT2); AIC(mchagas.occurrence_FSnoIT2)  # Display summary and AIC for the polynomial model

# Model for forest edge density (for_ED)
mchagas.occurrence_ED <- gam(chagas.occurrence ~ scale(for_ED) +offset(IDH) + s(year, bs = "re") + s(X, Y), family = binomial(link = "logit"), data = chagas.occurrence_data)
#summary(mchagas.occurrence_ED); AIC(mchagas.occurrence_ED)  # Display summary and AIC for for_ED model

# Model for patch density (for_PD)
mchagas.occurrence_PD <- gam(chagas.occurrence ~ scale(for_PD) + offset(IDH) + s(year, bs = "re") + s(X, Y), family = binomial(link = "logit"), data = chagas.occurrence_data)
#summary(mchagas.occurrence_PD); AIC(mchagas.occurrence_PD)  # Display summary and AIC for for_PD model
```

## Double additive Models I: indigenous territories and forest cover outside IT
Build additive models combining Indigenous Territories (tot_IT) and forest cover outside Indigenous Territories (FS_noIT)
```{r}
# Additive model with polynomial terms for tot_IT and FS_noIT
mchagas.occurrence_ITFSno <- gam(chagas.occurrence ~ scale(tot_IT) + poly(scale(FS_noIT), 2) + offset(IDH) + s(year, bs = "re") + s(X, Y), family = binomial(link = "logit"), data = chagas.occurrence_data)
# summary(mchagas.occurrence_ITFSno); AIC(mchagas.occurrence_ITFSno)  # Display summary and AIC for the model
```

### Double additive models II: overall forest cover and fragmentation
Build models combining overall forest cover (FS_PLAND) and fragmentation (for_ED and for_PD)
```{r}
# Additive model combining FS_PLAND and for_ED
mchagas.occurrence_FSED <- gam(chagas.occurrence ~  poly(scale(FS_PLAND), 2) + scale(for_ED) + offset(IDH) + s(year, bs = "re") + s(X, Y), family = binomial(link = "logit"), data = chagas.occurrence_data)
# summary(mchagas.occurrence_FSED); AIC(mchagas.occurrence_FSED)  # Display summary and AIC for the model

# Additive model combining FS_PLAND and for_PD
mchagas.occurrence_FSPD <- gam(chagas.occurrence ~ poly(scale(FS_PLAND), 2) + scale(for_PD) + offset(IDH) + s(year, bs = "re") + s(X, Y), family = binomial(link = "logit"), data = chagas.occurrence_data)
# summary(mchagas.occurrence_FSPD); AIC(mchagas.occurrence_FSPD)  # Display summary and AIC for the model
```

### Double additive models III: indigenous territories and fragmentation
```{r}
mchagas.occurrence_TIED <- gam(chagas.occurrence ~ scale(tot_IT) + scale(for_ED) + offset(IDH) + s(year, bs = "re") + s(X,Y), family= binomial(link = "logit"), data = chagas.occurrence_data)
# summary(mchagas.occurrence_TIED); AIC(mchagas.occurrence_TIED)

mchagas.occurrence_TIPD <- gam(chagas.occurrence ~ scale(tot_IT) + scale(for_PD) + offset(IDH) + s(year, bs = "re") + s(X,Y), family= binomial(link = "logit"), data = chagas.occurrence_data)
# summary(mchagas.occurrence_TIPD); AIC(mchagas.occurrence_TIPD)
```

### Tripple additives: indigenous territories + forest cover outside IT + fragmentation
```{r}
mchagas.occurrence_ITFSED <- gam(chagas.occurrence ~ scale(tot_IT) +  poly(scale(FS_noIT),2) + scale(for_ED) + offset(IDH) + s(year, bs = "re") + s(X,Y), family= binomial(link = "logit"), data = chagas.occurrence_data)
# summary(mchagas.occurrence_ITFSED); AIC(mchagas.occurrence_ITFSED)

mchagas.occurrence_ITFSPD <- gam(chagas.occurrence ~ scale(tot_IT) + poly(scale(FS_noIT),2) + scale(for_PD) + offset(IDH) + s(year, bs = "re") + s(X,Y), family= binomial(link = "logit"), data = chagas.occurrence_data)
# summary(mchagas.occurrence_ITFSPD); AIC(mchagas.occurrence_ITFSPD) 
```

## Interactive Models I: indigenous territories * forest cover outside IT
```{r}
# interativos IT * forest (FS_noIT)
#mchagas.occurrence_ITFSno1i <- gam(chagas.occurrence ~ poly(scale(tot_IT),2) * scale(FS_noIT) + offset(IDH) + s(year, bs = "re") + s(X,Y), family= binomial(link = "logit"), data = #chagas.occurrence_data) # TO DELETE

mchagas.occurrence_IT1FSnoi <- gam(chagas.occurrence ~ scale(tot_IT) * poly(scale(FS_noIT),2) + offset(IDH) + s(year, bs = "re") + s(X,Y), family= binomial(link = "logit"), data = chagas.occurrence_data)
# summary(mchagas.occurrence_IT1FSnoi); AIC(mchagas.occurrence_IT1FSnoi) 
```

### Interactive Models II: overall forest cover * fragmentation
```{r}
## interactive landscape structure
mchagas.occurrence_FSEDi <- gam(chagas.occurrence ~ poly(scale(FS_PLAND),2) * scale(for_ED) + offset(IDH) + s(year, bs = "re") + s(X,Y), family= binomial(link = "logit"), data = chagas.occurrence_data)
# summary(mchagas.occurrence_FSEDi); AIC(mchagas.occurrence_FSEDi) 

mchagas.occurrence_FSPDi <- gam(chagas.occurrence ~ poly(scale(FS_PLAND),2) * scale(for_PD) + offset(IDH) + s(year, bs = "re") + s(X,Y), family= binomial(link = "logit"), data = chagas.occurrence_data)
# summary(mchagas.occurrence_FSPDi); AIC(mchagas.occurrence_FSPDi) 
```

## Interactive Models III: indigenous territories * fragmentation
```{r}
mchagas.occurrence_TIEDi <- gam(chagas.occurrence ~ scale(tot_IT) * scale(for_ED) + offset(IDH) + s(year, bs = "re") + s(X,Y), family= binomial(link = "logit"), data = chagas.occurrence_data)
# summary(mchagas.occurrence_TIEDi); AIC(mchagas.occurrence_TIEDi) 

mchagas.occurrence_TIPDi <- gam(chagas.occurrence ~  scale(tot_IT) * scale(for_PD) + offset(IDH) + s(year, bs = "re") + s(X,Y),  family= binomial(link = "logit"), data = chagas.occurrence_data)
# summary(mchagas.occurrence_TIPDi); AIC(mchagas.occurrence_TIPDi) 
```

### Additive and interative model but keeping interaction with only two varibles: 

#### indigenous territories + ( forest cover outside IT * fragmentation )
```{r}
mchagas.occurrence_ITiFSED <- gam(chagas.occurrence ~ scale(tot_IT) + poly(scale(FS_noIT),2) * scale(for_ED) + offset(IDH) + s(year, bs = "re") + s(X,Y), family= binomial(link = "logit"), data = chagas.occurrence_data)
# summary(mchagas.occurrence_ITiFSED); AIC(mchagas.occurrence_ITiFSED)

mchagas.occurrence_ITiFSPD <- gam(chagas.occurrence ~ scale(tot_IT) + poly(scale(FS_noIT),2) * scale(for_PD) + offset(IDH) + s(year, bs = "re") + s(X,Y), family= binomial(link = "logit"), data = chagas.occurrence_data)
# summary(mchagas.occurrence_ITiFSPD); AIC(mchagas.occurrence_ITiFSPD)
```

#### forest cover outside IT + ( indigenous territories * fragmentation )
```{r}
mchagas.occurrence_FSiTIED <- gam(chagas.occurrence ~  poly(scale(FS_noIT),2) + scale(tot_IT) * scale(for_ED) + offset(IDH) + s(year, bs = "re") + s(X,Y), family= binomial(link = "logit"), data = chagas.occurrence_data)
# summary(mzoonotic_FSiTIED); AIC(mchagas.occurrence_FSiTIED)

mchagas.occurrence_FSiTIPD <- gam(chagas.occurrence ~ poly(scale(FS_noIT),2) + scale(tot_IT) * scale(for_PD) + offset(IDH) + s(year, bs = "re") + s(X,Y), family= binomial(link = "logit"), data = chagas.occurrence_data)
# summary(mchagas.occurrence_FSiTIPD); AIC(mchagas.occurrence_FSiTIPD)
```

#### fragmentation + ( indigenous territories * forest cover outside IT )
```{r}
mchagas.occurrence_EDiFSTI1 <- gam(chagas.occurrence ~ poly(scale(FS_noIT),2) * scale(tot_IT) + scale(for_ED) + offset(IDH) + s(year, bs = "re") + s(X,Y), family= binomial(link = "logit"), data = chagas.occurrence_data)
summary(mchagas.occurrence_EDiFSTI1); AIC(mchagas.occurrence_EDiFSTI1)

#mchagas.occurrence_EDiFS1TI <- gam(chagas.occurrence ~ scale(FS_noIT) * poly(scale(tot_IT),2) + scale(for_ED) + offset(IDH) + s(year, bs = "re") + s(X,Y), family= binomial(link = "logit"), data = chagas.occurrence_data)
#summary(mchagas.occurrence_EDiFS1TI); AIC(mchagas.occurrence_EDiFS1TI) TO DELETE

mchagas.occurrence_PDiFSTI1 <- gam(chagas.occurrence ~ poly(scale(FS_noIT),2) * scale(tot_IT) + scale(for_PD) + offset(IDH) + s(year, bs = "re") + s(X,Y), family= binomial(link = "logit"), data = chagas.occurrence_data)
# summary(mchagas.occurrence_PDiFSTI1); AIC(mchagas.occurrence_PDiFSTI1)

#mchagas.occurrence_PDiFS1TI <- gam(chagas.occurrence ~ scale(FS_noIT) * poly(scale(tot_IT),2) + scale(for_PD) + offset(IDH) + s(year, bs = "re") + s(X,Y), family= binomial(link = "logit"), data = chagas.occurrence_data)
# summary(mchagas.occurrence_PDiFS1TI); AIC(mchagas.occurrence_PDiFS1TI)
```

# Model Selection Using AIC Comparison
Perform model selection by comparing all models based on AIC using AICtab function
```{r}
(mod.sel <- AICtab(
              # Absence of effect (aka Baseline Model, 'null')
              mchagas.occurrence0,
              # Single-Predictor Models
              mchagas.occurrence_FS2,
              mchagas.occurrence_IT,
              mchagas.occurrence_FSnoIT2,
              mchagas.occurrence_ED,
              mchagas.occurrence_PD,
              # Double additive Models I: indigenous territories and forest cover outside IT
              mchagas.occurrence_ITFSno,
              # Double additive models II: overall forest cover and fragmentation
              mchagas.occurrence_FSED,
              mchagas.occurrence_FSPD,
              # Double additive models III: indigenous territories and fragmentation
              mchagas.occurrence_TIED,
              mchagas.occurrence_TIPD,
              # Tripple additives: indigenous territories + forest cover outside IT + fragmentation
              mchagas.occurrence_ITFSED,
              mchagas.occurrence_ITFSPD,
              ## Interactive Models I: indigenous territories * forest cover outside IT
              mchagas.occurrence_IT1FSnoi,
              # Interactive Models II: overall forest cover * fragmentation
              mchagas.occurrence_FSEDi,
              mchagas.occurrence_FSPDi,
              # Interactive Models III: indigenous territories * fragmentation
              mchagas.occurrence_TIEDi,
              mchagas.occurrence_TIPDi,
              # Additive and interative model but keeping interaction with only two varibles: 
              # indigenous territories + ( forest cover outside IT * fragmentation )
              mchagas.occurrence_ITiFSED,
              mchagas.occurrence_ITiFSPD,
              # forest cover outside IT + ( indigenous territories * fragmentation )
              mchagas.occurrence_FSiTIED,
              mchagas.occurrence_FSiTIPD,
              # fragmentation + ( indigenous territories * forest cover outside IT )
              mchagas.occurrence_EDiFSTI1,
              mchagas.occurrence_PDiFSTI1,
       base = TRUE, weights = TRUE))  # Include null model and calculate weights
```

# Top selected model output # note that 2 models were similarly plausible but I will proceed with the 'first' one 

Summarize the selected model to examine the fitted results
```{r}
summary(mchagas.occurrence_ITiFSED)
```

Tidy up the model summary to get a clean output of the parametric terms
```{r}
tidy(mchagas.occurrence_ITiFSED, parametric = TRUE)  #
```

# Model Diagnostics

## Dispersion

Test for overdispersion in the residuals using DHARMa's testDispersion()
```{r}
DHARMa::testDispersion(mchagas.occurrence_ITiFSED)
``` 

## Test and plot residuals

Generate residual diagnostic plots to visually inspect residuals and validate model assumptions. Creates diagnostic plots such as residual vs. fitted plots, QQ plots, and others for more in-depth inspection.
```{r}
testResiduals(mchagas.occurrence_ITiFSED, plot = TRUE)
```

Rarer diseases were not plotted in study